## Tbx5 gene expression analysis
### Introduction

This part of the work provides the analysis that was performed using the
data from eight ChIP-seq experiments which analyzed gene expression of 
*Tbx5* transcription factor in *Mus musculus* organism.\

The experiments that were used for *Mus musculus* *Tbx5* gene expression
analysis are listed below.\

***Mus musculus:***\

 - [1. EXP058852 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP058852)(*heart ventricle*)\
 - [2. EXP030898 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP030898)(*HL-1 (cardiac muscle cells)*)\
 - [3. EXP062056 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP062056)(*neonatal cardiac fibroblast cells line expressing large T antigen*)\
 - [4. EXP058843 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP058843)(*mouse embryonic fibroblasts*)\
 - [5. EXP058847 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP058847)(*mouse embryonic fibroblasts*)\
 - [6. EXP058850 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP058850)(*mouse embryonic fibroblasts*)\
 - [7. EXP058856 ](https://gtrd.biouml.org/#!table/gtrd_current.experiments/Details/ID=EXP058856)(*mouse embryonic fibroblasts*)\

The analysis provides the following information:\
 **1.** Total peak counts of each performed experiment;\
 **2.** Peak count distribution within different chromosomes.\

```{r Libraries}
library(pacman)
p_load(data.table, rtracklayer, randomcoloR, ggplot2, ggthemes,
       plyranges, ggpubr, BRGenomics, reshape2, plotly, heatmaply,
       dplyr, gplots, genomation, ggseqlogo, BSgenome.Mmusculus.UCSC.mm10,
       BSgenome, Biostrings, scales)
```

### Total peak counts
The barplot below shows how many *Tbx5* gene peaks were detected while analyzing
four different types of *Mus musculus* cells.

```{r fig, fig.width = 4, fig.height = 7, fig.align = "center"}
# Creating Sample Key table:
samples <- read.csv(file = "/home/daniele/Desktop/III_course/II_semester/Kursinis_darbas/TF_analysis/Generated_files/sample_key.csv")

# Importing BigBed format files:
pth <- "../Inputs/BigBed/"
bbfiles <- list.files(path = pth, "*bb")
grl <- GRangesList()
chr_abr <- c(paste0("chr", 1:19), "chrX", "chrY")

for (i in 1:length(bbfiles)) {
    grl[[i]] <- import(paste0(pth, bbfiles[i])) %>%
                       filter(seqnames %in% chr_abr)
    names(grl)[i] <- samples$Graph_names[samples$Filename == bbfiles[i]]
}

peaks <- lapply(grl, length) %>%
            as.data.table()  %>%
            melt() %>%
            as.data.frame()  %>%
            setnames(., c("Experiment", "Peak_count"))

col_fill <- randomColor(1, hue = "red", luminosity = "dark")

bplot1 <- ggplot(peaks, aes(x = Experiment, y = Peak_count)) +
                geom_bar(stat = "identity", position = "dodge", width = 0.5,
                         color = "black", fill = "#930d1f") +
                labs(x = "", y = "Pikų skaičius", size = 5) +
                ylim(0, 200000) +
                scale_y_continuous(labels = label_number(suffix = " K",
                                scale = 1e-3))

png(file = "../Figures/total_peak_counts.png", width = 400)

bplot1 + theme(
    panel.background = element_rect(fill = "#eeeef1", colour = "#4c0001"),
    panel.grid.major.y = element_line(colour = "#cab5b5", size = 0.3,
                                      linetype = "dashed"),
    panel.grid.minor.y = element_line(colour = "#cab5b5", size = 0.3,
                                      linetype = "dashed"),
    panel.grid.major.x = element_line(colour = "#cab5b5", size = 0.2,
                                      linetype = "longdash"),
    panel.grid.minor.x = element_line(colour = "#cab5b5", size = 0.2,
                                      linetype = "longdash"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11,
                               face = "bold", color = "black"),
    axis.text.y = element_text(size = 11, face = "bold", color = "black"),
    axis.title.x = element_text(size = 2),
    axis.title.y = element_text(size = 16),
    plot.title = element_text(hjust = 0.5, face = "bold")) +
    annotate("rect", xmin = 1.5, xmax = 4.5, ymin = -1, ymax = 170000,
             alpha = 0, color = "#ffae00", size = 1) +
    annotate("rect", xmin = 5.5, xmax = 6.5, ymin = -1, ymax = 25000,
             alpha = 0, color = "#ffae00", size = 1)

dev.off()

# Determining which experiment has a maximum peak count:
max_count <- peaks[which.max(peaks$Peak_count), ]
min_count <- peaks[which.min(peaks$Peak_count), ]
```

According to the barchart, the greatest number of *Tbx5* gene expression peaks was
detected in the cells that were analyzed in
***```r max_count```*** experiment.
The smallest number of *Tbx5* gene peaks was observed in the dataset that was
retrieved from ***```r min_count```*** experiment.

### Peak counts by chromosome

The barcharts below show how peak counts are distributed within
different chromosomes. Each bar represents separate experiment.

```{r fig2, fig2.align = "center"}
peak_counts <- data.frame(matrix(ncol = 3, nrow = 0)) %>%
                    setNames(., c("Name", "Chromosome", "Peak_count"))

count_peaks <- function(name, objects) {
    for (chr in chr_abr) {
        peaks <- objects[[name]] %>%
                    filter(seqnames == chr) %>%
                    length()
        peak_counts[nrow(peak_counts) + 1,] = c(name, chr, peaks)
    }
    return(peak_counts)
}

peak_counts <- lapply(names(grl), count_peaks, objects = grl) %>% bind_rows()
peak_counts


cas <- unique(peak_counts$Chromosome)
peak_counts$Chromosome <- factor(peak_counts$Chromosome, levels = c(cas))
peak_counts$Name <- factor(peak_counts$Name, levels = unique(peak_counts$Name))
print(peak_counts)

col_fill <- randomColor(1, hue = "red", luminosity = "dark")

# Pastaba: būtina pridėti 'as.numeric', nes kitu atveju Peak_count yra
# traktuojamas kaip 'string' - stulpelių dydžiai atvaizduojami
# leksikografiškai - gaunamas klaidingas atvaizdavimas (pvz. 67
# reikšmę atvaizduojantis stulpelis yra daug didesnis už
# stulpelį, kuris atvaizduoja 2948)!

bplot2 <- ggplot(peak_counts, aes(x = Name, y = as.numeric(Peak_count))) +
            geom_bar(stat = "identity", color = "black", fill = col_fill) +
            facet_wrap(~ Chromosome, ncol = 7) +
            theme(legend.position = "none") +
            ylab("Pikų skaičius") +
            xlab("") +
            scale_y_continuous(labels = label_number(suffix = " K",
                               scale = 1e-3))

png(file = "../Figures/peak_counts_by_chr.png", width = 1200, height = 500)
    
bplot2 + theme_linedraw() +
theme(axis.text.x = element_text(angle = 90, size = 12, vjust = 0.5),
      axis.text.y = element_text(size = 14, face = "bold"),
      axis.title.x = element_text(size = 14, colour = "black"),
      axis.title.y = element_text(size = 20, colour = "black"),
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(colour = "black", face = "bold", size = 16))
dev.off()
```

### Peak overlaps between samples
The heatmap below helps to indicate the biggest number of peak overlaps between 7
experiments.\

```{r}
coef_matrix <- matrix(nrow = length(grl), ncol = length(grl))

jaccard <- function(a, b) {
    len <- reduce(c(grl[[a]], grl[[a]])) %>% length()
    return((length(GenomicRanges::intersect(grl[[a]], grl[[b]])) / len) * 100)
}

for (i in 1:length(grl)) {
    for (y in 1:length(grl)) {
        coef_matrix[i, y] = jaccard(i, y)
    }
}

colnames(coef_matrix) <- names(grl)
rownames(coef_matrix) <- names(grl)

coef_mat1 <- coef_matrix
coef_mat2 <- coef_matrix

coef_mat1[lower.tri(coef_mat1, diag = TRUE)] <- NA
coef_mat2[upper.tri(coef_mat2, diag = TRUE)] <- NA

coef_mat_bind = rbind(data.matrix(coef_mat1), data.matrix(coef_mat2))
melt_coef_mat <- melt(coef_mat_bind, na.rm = TRUE)

overlaps <- ggplot(melt_coef_mat, aes(x = Var2, y = Var1, fill = value)) +
                 geom_tile(color = "black") +
                 geom_text(aes(label = round(value, digits = 3)),
                           color = "#030101", size = 4.5) +
                 labs(x = "", y = "") +
                 scale_fill_gradient(low = "#ffee8e", high = "#ab1f1f") +
                 guides(fill = guide_colourbar(title = "Koeficientas",
                        face = "bold")) +
                 theme(axis.text = element_text(size = 12, colour = "black",
                       face = "bold"),
                       axis.text.x = element_text(angle = 45, hjust = 1,
                                                  vjust = 1),
                       axis.title.x = element_text(size = 14, colour = "black"),
                       axis.title.y = element_text(size = 14, colour = "black"),
                       panel.grid.major = element_line(color = "#eeeeee"),
                       plot.title = element_text(hjust = 0.5, size = 16,
                       face = "bold"),
                       legend.position="bottom")

png(file = "../Figures/peak_overlaps.png", width = 450)
overlaps
dev.off()
```


### Motif determination using PWM matrix
The following analysis is focused on analysing the sequences based on the
peaks and trying to determine Tbx5 motif hits in sequences using PWM matrix.

```{r}
library(ggseqlogo)
library(dplyr)
library(Biostrings)
library(rtracklayer)
PATH <- "/home/daniele/Desktop/III_course/II_semester/Kursinis_darbas/TF_analysis"
mpwm <- read.table(paste0(PATH, "/Generated_files/TBX5_MOUSE.H11MO.0.D.pwm"))
mpwm <- t(mpwm)
rownames(mpwm) <- c("A","C","G","T")
tbx5_motif <- ggseqlogo(mpwm)
png(file = paste0(PATH, "/Figures/tbx5_motif.png"), height = 320)
tbx5_motif
dev.off()

# Calculating Tbx5 hits within fetched fasta files of peaks:
pth_fasta <- paste0(PATH, "/Inputs/FASTA/")
peak_sequences <- list.files(path = pth_fasta, "*fasta")

hit_vec <- c()
tbx5_hits <- data.frame(matrix(ncol = 3, nrow = 0)) %>%
                setNames(., c("Sample", "Peak_count", "Tbx5_hits"))

# Function that calculates Tbx5 motif hits in each sample:
find_motif_hits <- function(sequences) {
    for (i in 1:length(sequences)) {
        hits <- countPWM(as.matrix(mpwm), sequences[[i]], min.score="75%")
        if (hits == 0) { next }
        else { hit_vec <- c(hit_vec, hits)
        print(matchPWM(as.matrix(mpwm), sequences[[i]])) }
    }
    return(sum(hit_vec))
}

# Function that calculates total peak count in each sample:
calculate_peaks <- function(filename) {
    name <- paste0(tools::file_path_sans_ext(filename), ".bb")
    file <- import(paste0("../Inputs/BigBed/", name)) %>%
                    filter(seqnames %in% chr_abr)
    region_count <- length(file)
    return(region_count)
}

# Creating a new dataframe that stores data:
#   - generated sample name (used in plots);
#   - Tbx5 motif hit count;
#   - Total peak count;
#   - Tbx5 motif percentage.

# TEMPORARY COMMENTED!
# columns <- c("Sample", "Motif_count", "Peak_count", "Percentage") 
# tbx5_motifs <- data.frame(matrix(nrow = 0, ncol = length(columns))) 
 
# colnames(tbx5_motifs) <- columns

# for (i in 1:length(peak_sequences)) {
#     filename <- peak_sequences[i]
#     bb_file <- paste0(tools::file_path_sans_ext(filename), ".bb")
#     samples_plot_name <- samples$Graph_names[samples$Filename == bb_file]

#     seq <- readDNAStringSet(paste0(pth_fasta, filename))
#     motif <- find_motif_hits(seq)
#     peaks <- calculate_peaks(filename)
#     percentage <- round((motif / peaks) * 100, 2)

#     data_row <- c(samples_plot_name, motif, peaks, paste0(percentage, "%"))
#     tbx5_motifs[nrow(tbx5_motifs) + 1, ] <- data_row
# }

# # write.csv(tbx5_motif, "tbx5_motif_data.csv")
# # # THE END OF THE COMMENT
library(reshape2)
library(ggplot2)

motif_data <- read.csv(file = paste0(PATH, "/Generated_files/tbx5_motif_data.csv"))
motif_data$Sample <- factor(motif_data$Sample, levels = motif_data$Sample)

subset_df <- motif_data[, 2:5]
melted_df <- melt(subset_df)
melted_df

tf_percent <- ggplot(data = melted_df, aes(x = Sample, y = value, fill = variable,
                                    label = value)) + 
                geom_bar(stat = "identity", colour = "#35170450", size = 0.5,
                         width = 0.8) +
                scale_fill_manual(values = c("#e3a15e", "#c7633b"),
                                  labels = c("Tbx5 motyvų skaičius",
                                             "Pikų skaičius")) +
                scale_y_continuous(labels = label_number(suffix = " K",
                               scale = 1e-3)) +
                guides(fill = guide_legend(title = "Spalvų paaiškinimas", size = 6)) +
                labs(title = "", x = "", y = "TF/Pikų skaičius") +
                theme(axis.text = element_text(size = 10, colour = "black"),
                      axis.text.x = element_text(angle = 45, hjust = 1,
                                                 vjust = 1, size = 12,
                                                 face = "bold"),
                      axis.text.y = element_text(size = 12, face = "bold"),
                      axis.title.x = element_text(size = 14, colour = "black"),
                      axis.title.y = element_text(size = 14, colour = "black"),
                      panel.grid.major = element_line(color = "#eeeeee"),
                      plot.title = element_text(hjust = 0.5, size = 16,
                                                face = "bold"),
                      panel.background = element_rect(fill = "#eeeef1",
                                                      colour = "#4c0001"),
                      panel.grid.major.y = element_line(colour = "#cab5b5",
                                                        size = 0.3,
                                                        linetype = "dashed"),
                      panel.grid.minor.y = element_line(colour = "#cab5b5",
                                                        size = 0.3,
                                                        linetype = "dashed"),
                      panel.grid.major.x = element_line(colour = "#cab5b5",
                                                        size = 0.2,
                                                        linetype = "longdash"),
                      panel.grid.minor.x = element_line(colour = "#cab5b5",
                                                         size = 0.2,
                                                         linetype = "longdash"),
                      legend.position = c(0.79, 0.9),
                      legend.title = element_text(size = 12), 
                      legend.text = element_text(size = 11))

png(file = paste0(PATH, "/Figures/tf_hit_percentage.png"), width = 400)
tf_percent
dev.off()
```

The bar chart below visualizes total identified motif count
for each sample.

```{r}
library(ggplot2)
identified_motifs <- read.csv(file = "../Generated_files/identified_motifs.csv")

motif_counts <- data.frame(Motif_count = colSums(!is.na(identified_motifs)))
rownames(motif_counts) <- NULL
motif_counts['Sample'] <- colnames(identified_motifs)
motif_counts <- motif_counts[, c(2, 1)]

motif_counts$Sample <- factor(motif_counts$Sample, levels = motif_counts$Sample)

motifs <- ggplot(motif_counts, aes(x = Sample, y = Motif_count)) +
                geom_bar(stat = "identity", fill = "#6b1313",
                            color = "#4a1004", width = .7) +
                labs(title = "", x = "",
                         y = "Nustatytų motyvų skaičius") +
                ylim(0, 400) +
                geom_text(aes(label = Motif_count), color = "#030101",
                              vjust = -1, size = 5) +
                theme(axis.text = element_text(size = 10, colour = "black"),
                      axis.text.x = element_text(angle = 45, hjust = 1,
                                                 vjust = 1, size = 12,
                                                 face = "bold"),
                      axis.text.y = element_text(size = 12, face = "bold"),
                      axis.title.x = element_text(size = 14, colour = "black"),
                      axis.title.y = element_text(size = 14, colour = "black"),
                      panel.grid.major = element_line(color = "#eeeeee"),
                      plot.title = element_text(hjust = 0.5, size = 16,
                                                face = "bold"),
                      panel.background = element_rect(fill = "#eeeef1",
                                                      colour = "#3a1010"),
                      panel.grid.major.y = element_line(colour = "#cab5b5",
                                                        size = 0.3,
                                                        linetype = "dashed"),
                      panel.grid.minor.y = element_line(colour = "#cab5b5",
                                                        size = 0.3,
                                                        linetype = "dashed"),
                      panel.grid.major.x = element_line(colour = "#cab5b5",
                                                        size = 0.2,
                                                        linetype = "longdash"),
                      panel.grid.minor.x = element_line(colour = "#cab5b5",
                                                        size = 0.2,
                                                        linetype = "longdash"),
                      legend.position = c(0.777, 0.788))

png(file = "../Figures/motifs_in_samples.png", width = 500)
motifs
dev.off()
```